---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paramounter

<!-- badges: start -->
<!-- badges: end -->

Peak picking software such as **xcms**, **MS-DIAL**, and **MZmine** requires the user
to supply processing parameters. Mass tolerance, noise threshold, expected peak width,
and minimum signal-to-noise ratio all have to be set before peak detection can run,
and the results depend heavily on the values chosen.

**paramounter** determines these values by measuring them from the raw data. It
examines individual chromatographic peaks in your files, records how each one behaves,
and summarises those measurements into processing parameters. It does not repeatedly
run a peak picker and score the results, so a single pass over the files is
sufficient. One file takes roughly twenty seconds.

The method was published by
[Guo and Huan (2022)](https://doi.org/10.1021/acs.analchem.1c04758), *Analytical
Chemistry* 94:4260. This package is an independent implementation. The original code
is at [HuanLab/Paramounter](https://github.com/HuanLab/Paramounter), and Jian Guo and
Tao Huan are credited as copyright holders.

## Installation

```r
# install.packages("pak")
pak::pak("wmoldham/paramounter")
```

Input files must be centroided. mzML, mzXML, and CDF are all supported.

## The method

The algorithm works on one narrow mass range at a time. Within such a range the data
are simple enough to separate noise from signal directly, which is what makes the
measurements possible.

````{=markdown}
```mermaid
flowchart TD
    RAW[/"centroided mzML · mzXML · CDF"/] --> READ["read_ms_data()"]
    READ --> BIN["bin_peaks()<br/>slice m/z into 0.05 Da bins"]

    subgraph PERBIN ["for each mass bin containing signal"]
        direction LR
        EIC["assemble_bin_traces()<br/>build the chromatogram"]
        SM["smooth_intensity()"]
        NOISE["estimate_noise()<br/>find this bin's noise floor"]
        ZOI["find_zoi()<br/>find peaks above the floor"]
        EIC --> SM --> NOISE --> ZOI
    end

    BIN --> EIC

    subgraph PERZOI ["for each peak found"]
        direction LR
        WALK["collect_zoi_masses()<br/>follow the mass trace"]
        FEAT["zoi_features()<br/>measure the peak"]
        WALK --> FEAT
    end

    ZOI --> WALK
    FEAT --> AGG["aggregate_files()<br/>pool, filter, and summarise"]
    AGG --> UP(["universal_parameters"])

    classDef io fill:#4a5568,stroke:#2d3748,color:#ffffff
    classDef step fill:#3182ce,stroke:#2c5282,color:#ffffff
    classDef out fill:#2f855a,stroke:#22543d,color:#ffffff
    class RAW io
    class READ,BIN,EIC,SM,NOISE,ZOI,WALK,FEAT,AGG step
    class UP out
    style PERBIN fill:#8888881a,stroke:#888888
    style PERZOI fill:#8888881a,stroke:#888888
```
````

### Step 1. Read the file

`read_ms_data()` reads one data file and returns the m/z values and intensities of
every MS1 scan, together with the retention time of each scan.

### Step 2. Divide the mass range into bins

`bin_peaks()` cuts the m/z axis into slices 0.05 Da wide and assigns every peak from
every scan to a slice. Bins containing no peaks are dropped. All later steps operate
on one bin at a time.

### Step 3. Build a chromatogram for the bin

`assemble_bin_traces()` takes one bin and returns the mean intensity within that bin
at each scan. This is an extracted-ion chromatogram. Scans with no peak in the bin
have an intensity of zero.

### Step 4. Smooth the chromatogram

`smooth_intensity()` applies a weighted moving average. The default half-window is
zero, which leaves the chromatogram unchanged. Increase it for noisy data.

### Step 5. Find the noise floor of the bin

`estimate_noise()` sorts the non-zero intensities from lowest to highest and steps
through them in blocks of ten. At each step it calculates the mean and standard
deviation of all intensities below that point. The scan stops as soon as the next
intensity is greater than the mean plus three standard deviations, which is taken as
the boundary between noise and signal. The intensity at that boundary is the noise
cutoff for the bin. The function returns the cutoff along with the mean and standard
deviation of the noise, both of which are needed later.

Each mass bin receives its own cutoff rather than one global threshold.

### Step 6. Find the peaks

`find_zoi()` locates every run of consecutive scans whose intensity exceeds the noise
cutoff, and identifies the highest point in each run. Each run is a candidate
chromatographic peak, referred to as a zone of interest.

### Step 7. Follow the mass trace of each peak

`collect_zoi_masses()` starts at the top of the peak and moves outward one scan at a
time in both directions. At each scan it records the m/z value closest to the m/z at
the top of the peak. It stops when the intensity falls below the noise cutoff, when a
scan contains no signal in the bin, or when the recorded m/z becomes an outlier by
the Tukey fence test, which indicates that a neighbouring peak has been reached. The
function returns the recorded m/z values and the first and last scan of the peak.

The recorded m/z values are repeated measurements of the same ion. Their spread is
the mass accuracy of the instrument for that ion.

### Step 8. Measure the peak

`zoi_features()` converts the output of step 7 into six numbers.

| measurement | how it is calculated |
|---|---|
| mass tolerance in ppm | `2 × sd(m/z values) / reference m/z × 1e6` |
| mass tolerance in Da | `2 × sd(m/z values)` |
| peak width in seconds | retention time of the last scan minus the first |
| peak width in scans | number of scans spanned |
| signal-to-noise ratio | apex intensity minus noise mean, divided by noise sd |
| peak height | intensity at the top of the peak |

A peak is discarded if its mass trace contains fewer than 2 or more than 199 values,
which removes single-point spikes and runs that have merged with a neighbour.

### Step 9. Mark peaks suitable for measuring drift

`flag_isolated_zoi()` marks peaks that have no neighbouring peak within 300 seconds
in the same mass bin. Only these are used in step 12, because a crowded peak cannot
be matched reliably between files. They can optionally be the only peaks the
tolerances are measured on as well; see
[When most zones are not peaks](#when-most-zones-are-not-peaks).

### Step 10. Discard unreliable peaks

Steps 3 to 9 repeat for every bin in every file, which typically yields several
thousand measured peaks per file. `aggregate_files()` pools them and applies a
threshold to the ppm measurements. Peaks whose mass trace is too scattered are noise
rather than signal, and would otherwise inflate the mass tolerance.

The threshold is taken from the pooled ppm values, at the 95th percentile of the
lowest 97 percent. It can also be set directly with `ppm_cutoff`.

### Step 11. Trim the threshold measurements

A threshold is summarised by its minimum, so a single unusual peak could determine
it. `aggregate_files()` discards the most extreme 3 percent from that end of each
threshold distribution (`noise`, `width_scans`, `sn`, `height`). The tolerance
distributions (`ppm`, `mz_diff`, `width_seconds`) are left whole, because they are
summarised at a quantile rather than at an extreme, and trimming first would compose
with it. Under `legacy = TRUE` every distribution is trimmed, as in the original.

### Step 12. Measure instrument drift

Given two or more files, `match_zoi_across_files()` pairs the isolated peaks from
step 9 across files, using a window of 0.015 Da and 30 seconds around each peak in
the first file. `estimate_instrument_shift()` then calculates, for each matched peak,
the range of its m/z and the range of its retention time across the files. These
describe how far the instrument drifts between injections.

### Output

`paramounter()` runs all of the above and returns a `universal_parameters` object.
It holds one distribution of measurements per quantity, together with the files and
settings used. The `summary` property tabulates them, and `plot()` draws a histogram
of each.

| quantity | what it describes |
|---|---|
| `ppm`, `mz_diff` | mass accuracy, relative and absolute |
| `noise` | noise floor across mass bins |
| `width_seconds`, `width_scans` | chromatographic peak width |
| `sn` | signal-to-noise ratio |
| `height` | peak intensity |
| `mass_shift`, `rt_shift` | instrument drift between files |

## Converting to software parameters

`to_xcms()`, `to_msdial()`, and `to_mzmine()` convert the measurements into settings.
Each distribution collapses to a single value taken from one end of it, according to
the role the parameter plays.

A tolerance is set from the upper end, so that it is wide enough to accommodate the
worst-behaved genuine peak. Those that govern how a peak is built are drawn from
`tolerance_quantile` rather than from the maximum, because a single noise-limited
measurement should not widen the tolerance for every peak in the file. Those that
govern grouping are still drawn from the maximum.

| measurement | xcms parameter | taken from |
|---|---|---|
| `ppm` | `ppm` | `tolerance_quantile` |
| `width_seconds` | upper bound of `peakwidth` | `tolerance_quantile` |
| `mass_shift` | `binSize` for grouping | the largest value |
| `rt_shift` | `bw` for grouping | the largest value |

Under `legacy = TRUE` every tolerance is the maximum, as in the original method.

A threshold is set from the minimum, so that it is low enough to retain the weakest
genuine peak.

| measurement | xcms parameter |
|---|---|
| smallest `noise` | `noise` and the intensity part of `prefilter` |
| smallest `width_scans` | the scan-count part of `prefilter` |
| smallest `sn` | `snthresh`, with a floor of 3 |
| smallest `height` | minimum peak height for MS-DIAL and MZmine |

The peak-width bounds are the smallest width rounded up plus 4, and the
`tolerance_quantile` of the widths rounded up plus 5. Under `legacy = TRUE` the upper
bound is the largest width, and the original's special case for broad peaks applies:
when the largest width exceeds 35 seconds and peaks are tall relative to their width,
the range becomes 0 to the largest width rounded up plus 7, halved. It is not applied
by default, because a lower bound of zero asks `CentWave` for peaks of no width and
the height-to-width constant it tests was calibrated on Bruker Q-TOF counts.

### Grouping parameters and retention time correction

`bw` and `binSize` carry a caveat the others do not. They are derived from `rt_shift` and
`mass_shift`, which are measured on the files as acquired, but they are consumed by
`groupChromPeaks()`, which in the usual xcms order runs *after* `adjustRtime()`. Retention
time correction exists to remove between-injection drift, so by the time grouping happens,
much of the drift these parameters were sized against is already gone.

The estimate is therefore conservative, and where alignment works well it can be
conservative enough to matter: grouping tolerates the full uncorrected drift, so peaks the
chromatography resolves may be collected into a single feature. In a plasma lipidomics run
of 216 injections, `bw` was measured at 7.2 seconds while residual drift after peak group
alignment was 4.8 seconds. At the measured value, 32% of features contained chromatographic
peaks spread over more than 20 seconds of retention time, and inspection of the extracted
ion chromatograms showed these to be co-eluting species merged together. Reducing `bw` to
just below the residual drift brought that to 8%, while the detection rate in pooled
quality control injections fell by under four percentage points.

If you align before grouping, treat the residual drift as a ceiling for `bw` rather than as
the answer, and sweep downwards from it. The negative mode of the same study had 9.6 seconds
of residual drift, but grouping at 8 seconds still merged 15% of features; the merge rate
only reached the positive mode's level at 6 seconds, and the quality control detection rate
was flat the whole way down. Detection falling off is the signal that a real feature is now
being split across injections, and it is a more direct guide than the drift figure alone.

The same argument applies to `binSize` for mass drift, though its effect is usually smaller.
What this package measures is the instrument as it ran; where a later processing step removes
part of what was measured, the parameter depending on it is worth revisiting.

## Usage

The package ships five urine data files, which the examples below use. Point
`list.files()` at your own directory to analyse your own data.

```r
library(paramounter)

files <- list.files(
  system.file("extdata", package = "paramounter"),
  pattern = "\\.mzXML$",
  full.names = TRUE
)

params <- paramounter(files)

params@summary
plot(params)
```

`to_xcms()` returns the three parameter objects an xcms workflow needs, ready to pass
to the corresponding functions.

```r
xp <- to_xcms(params)

data <- xcms::findChromPeaks(data, xp$chrom_peaks)
data <- xcms::groupChromPeaks(data, xp$group)
data <- xcms::adjustRtime(data, xp$retention)
```

Two or more files are required to measure instrument drift. With a single file the
`mass_shift` and `rt_shift` distributions are empty, `to_xcms()` issues a warning,
and the grouping parameters fall back to defaults.

`to_msdial()` and `to_mzmine()` return a table of parameter names and values, which
are entered into those programs manually.

### Settings

Every setting is held in a `pm_config` object, which is passed to `paramounter()`.
The numeric settings follow the published method.

```r
pm_config(
  bin_width = 0.05,          # width of the mass bins, in Da
  smooth_half_window = 0,    # smoothing applied to each chromatogram
  ppm_cutoff = NULL,         # NULL determines the cutoff from the data
  isolation_gap = 300,       # seconds of separation required by step 9
  match_mz_tol = 0.015,      # matching window used by step 12
  match_rt_tol = 30,
  trim = 0.97,               # kept from each threshold distribution
  tolerance_quantile = 0.95, # taken from each tolerance distribution
  tolerance_isolated = FALSE,# measure tolerances on isolated zones only
  tolerance_min_scans = 1,   # ...and only on zones at least this many scans wide
  legacy = FALSE
)
```

### Checking the parameters before you use them

Every parameter is drawn from one end of a measured distribution. When the
distribution is tight, that value is representative. When it has a long tail, the
value is set by the worst-measured peaks in the file rather than by what the
instrument can do.

`check_estimates()` shows the gap:

```r
check_estimates(params)
#>        quantity      kind  estimate    median ratio
#> 1           ppm tolerance 2.000e+01 5.664e+00  3.53
#> 2       mz_diff tolerance 1.000e-02 1.523e-03  6.56
#> 3 width_seconds tolerance 4.100e+01 5.851e+00  7.01
#> 6         noise threshold 2.980e+02 4.640e+02  1.56
#> 9            sn threshold 3.000e+00 1.329e+01  4.43
```

`ratio` is how many times further from the median the parameter sits. Read the
`kind` column before acting on a large one, because the two kinds fail
differently.

A **tolerance** decides how a peak is built. Setting it too loosely cannot be
undone later. On an Orbitrap dataset, a `ppm` of 46 instead of 10 found twice as
many chromatographic peaks and ended up with *fewer* features, because the extra
detections did not survive grouping. Lower `tolerance_quantile` if you see a
large ratio here.

A **threshold** decides how many candidates pass. Setting it loosely only adds
candidates, and your own quality control removes the ones that do not reproduce.
A large ratio here is the method working as intended. Tightening these has been
measured to discard real features, so leave them alone unless your own data says
otherwise.

### When most zones are not peaks

Lowering `tolerance_quantile` helps when a tolerance is set by a tail of poorly
measured peaks. It cannot help when the bulk of the distribution is poor, which
happens when most measured zones are not chromatographic peaks at all.

On a polarity-switching HILIC Orbitrap run, with 1.2 seconds between scans of one
polarity, the median zone was 2 scans wide and only 230 of 27,654 zones were isolated.
The 95th-percentile `ppm` was 45 in every intensity decile, so no quantile rescued it.
The upper peak-width bound came out at 27 seconds, below the median width of a real
peak on that column (about 25 seconds), so `CentWave` split broad peaks. A third of the
integrations of known metabolites came out at a fraction of their true area.

`tolerance_isolated` and `tolerance_min_scans` restrict the tolerance distributions to
zones that are chromatographic peaks: no neighbour in the mass bin within
`isolation_gap`, and at least that many scans wide. On the same file, isolated zones of
5 or more scans gave a 95th-percentile `ppm` of 2.7 and peak widths from 5 to 104
seconds. Both conditions are needed there: the isolated zones alone still had a
95th-percentile `ppm` of 30, because an isolated 2-scan blip is still a blip.

```r
params <- paramounter(files, pm_config(tolerance_isolated = TRUE, tolerance_min_scans = 5))
```

The thresholds still use every zone, following the rule above. The options are off by
default and are not a general improvement. On the Orbitrap lipidomics data behind 0.2.0
the default already gives the validated `ppm`, and turning them on would tighten it
from 8-10 to 5, a value never tested there. A zone-level `ppm` is also a narrower
quantity than what `CentWave` bounds, which is the scan-to-scan scatter along a whole
peak including its weak edges, so a very small value may need headroom before it is
used.

### Reproducing the original exactly

`legacy` controls whether the package applies its corrections to the original
implementation or reproduces it exactly.

| behaviour | `legacy = FALSE` (default) | `legacy = TRUE` |
|---|---|---|
| peak width in scans | counts the scans spanned | counts one scan too many |
| a peak counts as isolated when | both neighbours are distant | only the following peak is distant |
| matching peaks across files | takes the closest candidate | takes the first candidate in the window |
| mass and peak-width tolerances | a quantile of the distribution | the largest value measured |
| zones the tolerances are measured on | every zone, unless `tolerance_isolated` or `tolerance_min_scans` is set | every zone |
| very wide peaks | no special case | halves the upper bound, drops the lower bound to zero |
| grouping bandwidth for xcms | taken from the measured drift | fixed at 5 |

The default is `FALSE`: the corrections are applied unless you ask for the original.
Set `legacy = TRUE` when you want to compare against published results, as the next
section does.

## Reproducing the published results

The five files shipped with the package are the urine samples distributed with the
original tool. They were acquired on a Bruker maXis impact Q-TOF in positive mode with
reversed-phase chromatography, and each contains 633 MS1 scans.

The call below sets `legacy = TRUE`, since the point here is to match the original.
For a real analysis leave it at its default.

```{r run, message = FALSE, warning = FALSE}
library(paramounter)

files <- list.files(
  system.file("extdata", package = "paramounter"),
  pattern = "\\.mzXML$",
  full.names = TRUE
)

params <- paramounter(files, pm_config(ppm_cutoff = 30, legacy = TRUE))
params@summary
```

```{r xcms, message = FALSE, warning = FALSE}
to_xcms(params)$chrom_peaks
```

Eleven of the twelve values published for this dataset are reproduced exactly.

| parameter | published | paramounter |
|---|---|---|
| `ppm` | 30 | 30 |
| `peakwidth` | 0, 28.5 | 0, 28.5 |
| `mzdiff` | −0.01 | −0.01 |
| `snthresh` | 3 | 3 |
| `integrate` | 1 | 1 |
| `prefilter` | 3, 298 | 3, 298 |
| `noise` | 298 | 298 |
| `bw` | 5 | 5 |
| `minfrac` | 0.5 | 0.5 |
| `minsamp` | 1 | 1 |
| `max` | 100 | 100 |
| `mzwid` | 0.006 | 0.007 |

The ppm cutoff is supplied directly in the call above rather than determined from the
data. The original implementation requires the user to read this number from a plot
and type it into the second half of the analysis. The published value of 30 was
obtained that way. Left to determine the cutoff itself, this package calculates 32.4
and reports a `ppm` of 33.

The remaining difference concerns `mzwid`, the mass window used to group peaks
between samples. It is set from the largest instrument mass drift measured in step
12. One of the 95 matched peaks is paired incorrectly. A peak at m/z 481.12942 in the
first file is matched to a peak at 481.118316 in the fifth, 11 mDa away, although a
peak at 481.128420 lies 1 mDa away in the same window. The original takes the first
candidate found in the matching window rather than the closest, which is what
`legacy = TRUE` reproduces here. The incorrect pairing raises the measured drift from
0.0058 to 0.0071.

Correcting the matching rule on its own would give exactly the published 0.006, but
the default `legacy = FALSE` does not recover it: that setting also changes which
peaks qualify as isolated in step 9, which rebuilds the set of measured peaks and
lands back at 0.0071. So no single setting reproduces all twelve values at once.
Eleven of twelve under `legacy = TRUE` is the honest ceiling, and the reason for the
twelfth is understood. Why the original authors' own run avoided the mis-pairing is
not recoverable: it depends on the order of peaks within the matching window, which
their published output does not record.

## Citation

Please cite the original method.

> Guo J, Huan T. Mechanistic Understanding of the Discrepancies between Common Peak
> Picking Algorithms in Liquid Chromatography–Mass Spectrometry-Based Metabolomics.
> *Analytical Chemistry* 2022, 94(13), 4260–4268.
